---
title: "data table beyond the basics"
author: "Mark Hatcher"
date: "20/02/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Beyond the basics

### Tutorial source
This is a [link](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) to the tutorial source.

Set up the environment

```{r message = FALSE}
rm(list = ls())
options(stringsAsFactors = FALSE)
library(data.table)
library(dplyr)
```

Load in the flights and print out the top 5 and bottom five rows

```{r}
flights <- fread("data//flights14.csv")
flights %>% head()
```

## Basic groupby

```{r}
flights[, .(.N), by = .(origin)]
```

### Grouping by more than one column

```{r}
flights[, .(.N), by = c("origin", "dest")] %>% head()
```

We can filter and we can name the aggregated column

```{r}
flights[carrier == "AA", .(count = .N ), by = origin]
```

### Sorting groups using keyby
Data.table retains the original order of groups. This is intentional
and by design.  But at times we would automatically like to sort
by the variables that we grouped by.  "keyby" lets us sort by
groups **Use keyby to sort groups instead of key**

```{r}
flights[carrier == "AA", .(delay = mean(arr_delay)), 
        keyby = .(origin, dest)] %>% head()
```

## Chaining
We can keep tacking on expressions.....[][][][]

```{r}
flights[carrier == "AA", .(count = .N), by = .(origin, dest)][order(-count)][1:4]
```

## Expressions in By
In the following we are grouping on logical expressions and then counting the result

```{r}
flights[, .N, .(dep_delay > 0, arr_delay > 0)]

flights[, .N, by = .(dep_delay > 0, arr_delay > 0)]
```

## Special symbol .SD

```{r}
# set up a data.frame for the following
vct_id <- c("b", "b", "b", "a", "a", "c")
vct_a <- c(1, 2, 3, 4, 5, 6)
vct_b <- c(7, 8, 9, 10, 11, 12)
vct_c <- c(13, 14, 15, 16, 17, 18)
DT <- data.table(ID = vct_id, a = vct_a, b = vct_b, c = vct_c)
DT
```

## The mighty .SD

.SD stands for subset of data.  It by itself is a data.frame that holds the data for the current group defined using "by"

```{r}
# This returns things in the same order in which they were defined
DT[, print(.SD), by = ID]
```

This enables us to apply functions to columns.

Above we group by ID and this results in three groups.  Now what if we wanted to find the mean of each group
and do this for all columns.

There are 3 groups. And 3 columns. So there will be 9 means created.

```{r}
DT[, lapply(.SD, mean), by = ID]
```

### How can we specify the columns that we computer the mean on



```{r}
# 200 groups
flights[carrier == "AA", .N, by = .(origin, dest, month)] %>% nrow()

# have a look at the output here
flights[1:100, print(.SD), by = .(origin, dest, month), 
        .SDcols = c("arr_delay", "dep_delay")]

# we know there are 200 groups.  And there are two columns we create summary
# information for.  That makes it 200 rows by 5 columns
flights[carrier == "AA",
  lapply(.SD, mean),
  by = .(origin, dest, month),
  .SDcols = c("arr_delay", "dep_delay")] %>% head()

flights[carrier == "AA",
  lapply(.SD, mean),
  by = .(origin, dest, month),
  .SDcols = c("arr_delay", "dep_delay")] %>% dim()

```

### Show the first N rows of a group
```{r}
# show the first two entries for each month
flights[, head(.SD, 2), by = month] %>% head(6)
```

### Concatenating columns
In this example, for the index b we have the first three rows of **a** being 1, 2, 3.
And, we have the first few rows of **b** being 7,8 9

```{r}
# create a column called val.
# group by ID and then concatenate...
# basically doubles the number of rows
DT[, .(val = c(a,b)), by = ID] %>% head()


# DTa <- DT[, .(ID, val = c(a,b))


```

In the following, we cast the result to be list
For the column val, each row is **actually a list**
There will be three rows -- a row for each value in the group


```{r}
# this shows how its done
DT[, print(c(a,b)), by = ID]
```













