---
title: "data_table_basics"
author: "Mark Hatcher"
date: "20/02/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data table basics

### Tutorial source
This is a [link](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) to the tutorial source.

Set up the environment

```{r chunk_name_mark, echo=FALSE, message=FALSE}
rm(list = ls())
options(stringsAsFactors = FALSE)
library(data.table)
library(dplyr)
```

Load in the flights and print out the top 5 and bottom five rows

```{r}
flights <- fread("data//flights14.csv")
flights %>% head()
```

### Strings are not converted to factors

```{r}
DT = data.table(ID = c("b","b","b","a","a","c"), a = 1:6, b = 7:12, c = 13:18)
DT$ID %>% class()
# print the number of rows to display
getOption("datatable.print.nrows")
```

## Basic syntax
This is row, column and grouping...

### Filtering
```{r}
flights[origin == "JFK" & month == 6L]
flights[1:2]
```

### Selecting columns

```{r}
# this returns a vector
flights[, arr_delay] %>% class()

# this returns a data.table
flights[, list(arr_delay)] %>% class()

# this is a synonym to a list
flights[, .(arr_delay, dep_delay)] %>% head()

# this is how we can change names
flights[, .(arrival = arr_delay, departure = dep_delay)] %>% head()
```

### Column Expressions
```{r}
# how many trips had a total delay that was minus  (arrived before or departed before)
# prints a single number
flights[, sum((arr_delay + dep_delay) < 0)]

# what is the average delay for JFK origin flights in June
flights[origin == "JFK" & month == 6L,  .(delay = mean(dep_delay))]
```

## Data table can optimise
Because the three i, j and by are inside [...] data.table can optimise
the query before evaluation...rather than each separately...


### Counting things -- the .N operator

```{r}
flights[origin == "JFK" & month == 6L, .N]
```

### Using the data.frame syntax to select columns

```{r}
# the following with = FALSE allows us to use data.frame syntax
flights[1:4, c("arr_delay", "dep_delay"), with = FALSE]

# this gives a different result - I cant see any difference.
flights[1:4, c("arr_delay", "dep_delay"), with = TRUE]
```
with = TRUE is default in data.table because we can  do much more by allowing j to handle expressions

### Ordering things
```{r}
flights[order(origin, -dest)] %>% head()
```




